Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-04 Manuscript/MPESubmission/Revision/response.tex

% -----------------
% Specific Comment 59
% -----------------
\item{\textcolor{blue}{\textbf{Appendix C:} Tables 1-3 repeats 'of the'. Should label the first two rows as 'Best maximum likelihood tree' and 'Best Bayesian inference tree'}}

We removed the repetition in the caption. %Not sure I understand the second part of the comment.

% -----------------
% Specific Comment 60
% -----------------
\item{\textcolor{blue}{\textbf{Appendix D:} I find the online code repository thorough, but the repository is loaded with many files, especially for graphing. If it is possible to condense these files it may be easier for a reader to duplicate the study.}}

We updated the readme page of the github repository linking to the various R codes.
We also now provided a version of the analysis in R ran thoughout the manuscript as a step by step tutorial for a subset of data.
See \url{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data/blob/master/Analysis/Analysis_Demo.R} and Appendix D.

### 2015-08-04 Manuscript/MPESubmission/Revision/response.tex

\begin{enumerate}[(1)]
\item The Total Evidence methods allows to combine living and fossil data.
\item This study test the effect of missing morphological data on topology.
\item The number of living taxa with data as the most effect on recovering topology.
\item Bayesian inference outperforms Maximum Likelihood for correct clade recovery.
\item We suggest that the sampling effort should be focused on data for living taxa.
\end{enumerate}

% -----------------
% Comment 4
% -----------------
\item{\textcolor{blue}{\textbf{Introduction:} Needs stronger emphasis on what specifically is unique about this study compared to what the many other studies on missing data have done and found.
Clearly outlining the questions in this study that are left open from previous research is important because some of the findings in this study suggest that there may be more error associated with missing data than previous research has suggested.}}

We emphasised on novelty throughout the introduction by more clearly stating the difference between the present study and the literature and more clearly outlining the question of this study.

"Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework (i.e. using combined molecular and morphological data; e.g. Wiens et al., 2005; Manos et al., 2007; Pattinson et al., 2014).
For example, Wiens et al. (2005) only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.
Furthermore, Pattinson et al. (2014) assess the effect of removing both molecular and morphological data (“artificial extinction”) on the topology.
As for morphological and molecular data separately, these studies shows that missing data is not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.
These studies, however, are limited by their empirical-only approach (Wiens et al., 2005) and the scheme of generating missing data (Manos et al., 2007; Pattinson et al., 2014).
Also both studies focus on the palaeontological aspect of the question (i.e. the fossil data)." lines 85-98 @@@

@@

% -----------------
% Comment 14
% -----------------
\item{\textcolor{blue}{\textbf{Table caption:} Table caption should include a line about 'pooling across missing data schemes' to clarify what is presented.}}

%TG: WUT? Didn't understood that one at all.

% -----------------
% Comment 15
% -----------------
\item{\textcolor{blue}{\textbf{Previous research:} The missing-data problem has a long history in the literature and there are references to some of those older and recent papers.
But I feel it is necessary to summarize the following questions: what did those studies find, what are the questions those studies left open, and how does this study address them?}}

We emphasized more on the recent findings from the missing-data papers in the introduction, and discussion (see comment 4 and 10 above).
We also included that our results for the $M_F$ parameter for the Normalised Robinson-Foulds metric where in accordance with Manos et al. 2007 in the discussion about the $M_F$ parameter:

"These results are in agreement with Manos et al. (2007) where as few as 16 characters where sufficient for correctly assigning ``artificial fossils'' to their ``correct'' clade." lines @@@

### 2015-08-04 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
added Manos cites + discussion sentence

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework \protect\citep[\hl{i.e. using combined molecular and morphological data}; e.g.][]{Wiens01102005,pattinsonphylogeny2014}.}
\hl{For example, \protect\cite{Wiens01102005} }\hl{only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.}
\hl{Furthermore, \protect\cite{pattinsonphylogeny2014} }\hl{assess the effect of removing both molecular and morphological data (``artificial extinction'') on the topology.}
\hl{As for morphological and molecular data separately, these studies shows that missing data is not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.}
\hl{These studies, however, are limited by their empirical-only approach \protect\citep{Wiens01102005} }\hl{and the scheme of generating missing data \protect\citep{pattinsonphylogeny2014}.}
\hl{Also both studies focus on the palaeontological aspect of the question (i.e. the fossil data).}

@@

\subsubsection{Missing data for fossil taxa ($M_{F}$)}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this also reduces the probability of morphological characters for fossil taxa overlapping with the ones for living taxa.
This can lead to difficulties for the placement of certain taxa in the tree.
It is important, however, to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets metric) with increasing missing data in this parameter, clade conservation (i.e. Normalised Robinson-Foulds metric) is still relatively good (mode = 0.72) when the proportion of missing data is high ($M_{F}$ = 75\%).

>>>>

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework \protect\citep[\hl{i.e. using combined molecular and morphological data}; e.g.][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.}
\hl{For example, \protect\cite{Wiens01102005} }\hl{only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.}
\hl{Furthermore, \protect\cite{pattinsonphylogeny2014} }\hl{assess the effect of removing both molecular and morphological data (``artificial extinction'') on the topology.}
\hl{As for morphological and molecular data separately, these studies shows that missing data is not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.}
\hl{These studies, however, are limited by their empirical-only approach \protect\citep{Wiens01102005} }\hl{and the scheme of generating missing data \protect\citep{manosphylogeny2007,pattinsonphylogeny2014}.}
\hl{Also both studies focus on the palaeontological aspect of the question (i.e. the fossil data).}

@@

\subsubsection{Missing data for fossil taxa ($M_{F}$)}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this also reduces the probability of morphological characters for fossil taxa overlapping with the ones for living taxa.
This can lead to difficulties for the placement of certain taxa in the tree.
It is important, however, to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets metric) with increasing missing data in this parameter, clade conservation (i.e. Normalised Robinson-Foulds metric) is still relatively good (mode = 0.72) when the proportion of missing data is high ($M_{F}$ = 75\%).
\hl{These results are in agreement with \cite{manosphylogeny2007} where as few as 16 characters where sufficient for correctly assigning ``artificial fossils'' to their ``correct'' clade.}
